Wait-free Queues with Polylogarithmic Step Complexity

### 2022-09-29 thesis/thesis.tex
fix citations, review index deq

\pagebreak
\section{Our Queue}
Jayanti and Petrovic introduced a wait-free polylogarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We benefit from some ideas of their work to design a a polylogarithmic multi-enqueuer multi-dequeuer queue. Our algorithm despite them does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent operations. I our model there are $p$ procesess doing \nf{Enqueue(),Dequeue()} operations concurrently. We use a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by processes. Each process has a leaf which the order of operations invoked by the process is stored in it. When a process wishes to do an operation it appends the operation to its leaf and then tries to propagate its new operation up to the tree's root. In each node the ordering of operations propagated up to it is stored. All processes agree on the sequence stored in the root and it is defined to be the linearization ordering. 
\begin{figure}[hbt]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption{ّ\label{fig::blocktree}Each process has a leaf and in each node there is an ordering of operations stored. Each node tries to propagate its operations up to the root, which stores the total ordering of all operations.}  
\end{figure}

@@

%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
\paragraph{}
Now, we describe how to use the tree to implement a queue. Consider the following execution of operations. \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. What should a \texttt{Dequeue()} return? To compute the response of a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.

@@

The next lemma is used to prove the subblocks of two blocks in a node are disjoint.
\begin{lemma} \label{lem::headProgress}
 If $n\nf{.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
\end{lemma}
\begin{proof}
Consider the block $b$ written into \nf{n.blocks[$i$]} by \nf{CAS} at Line \ref{cas}. Block $b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}. Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value \nf{b}$^\prime$ by Invariant \ref{lem::headPosition}. Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked. The value written into \nf{$b$.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock($i$)}  was one less than the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}. Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($i-1$)}. By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $b^\prime\nf{.end\sub{left}} \leq \nf{$b$.end\sub{left}}$. The proof for \nf{end\sub{right}} is similar.
\end{proof}

@@

Now we can define the operations of a block using the definition of subblocks.
\begin{definition}[Operations of a block]\label{def::ops}
A block $b$ in a leaf represents an \nf{Enqueue()} if $b\nf{.element}\neq\nf{null}$. Otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue()}. The set of operations of block \nf{b} is the union of the operations in leaf subblocks of $b$. We denote the set of operations of block $b$ by ${ops(b)}$ and the union  of operations of a set of blocks $B$ by $ops(B)$.  We also say ${b}$ contains ${op}$ if $op\in{ops(b)}$.
\end{definition}

@@

\subsection{Correctness of GetEnqueue and IndexDequeue}

Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnqueue()} and \nf{IndexDequeue()}.

@@

For the induction step we prove if \nf{$n$.child.GetEnqueue($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$.  From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}). \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}. By Definitions \ref{def::ops}, \ref{def::subblock} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
\begin{align*}
&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \cup \\
&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
 \end{align*}
First we consider the case where the \nf{Enqueue} we are looking is in the left child. There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

@@

\begin{definition}\label{orderRefresh}  A Refresh is \it{successful} if it performs a successful \nf{CAS} on Line \ref{cas}. If \nf{Refresh} instance \nf{R\sub{1}} does its \nf{CAS} at Line \ref{cas} sooner than \nf{Refresh} instance \nf{R\sub{2}} we say \nf{R\sub{1}}  \it{happened} before \nf{R\sub{2}}.
\end{definition}

 Let \nf{i} be the value \nf{R\sub{n}}, a successful instance of \nf{Refresh()} on the node \nf{n}, reads from \nf{n.head}. \nf{R\sub{n}} does a successful \nf{CAS(null, new)} into \nf{n.blocks[h]}, where \nf{b} is the \nf{new}. Without loss of generality for the rest of this section assume \nf{n} is the left child of \nf{n.parent}. From now on we say \nf{p} as an abbreviation for \nf{n.parent}. Let \nf{R\sub{p}} be the first successful \nf{p.Refresh()} that reads some value greater than \nf{i} for \nf{left.head} and is contains \nf{b} in its created block \nf{s} in Line \ref{invokeCreateBlock}. From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh()} propagating \nf{b}. \nf{R\sub{p}} does a successful \nf{CAS(null, new)} into \nf{p.blocks[j]}, where \nf{s} is the \nf{new}.

Although other fields of \nf{b} are set while creating it, because the index of the superblock of \nf{b} is not known until it is propagated, \nf{R\sub{n}} cannot set the \nf{super} field of a \nf{b} while creating it. One approach is to set the \nf{super} field of \nf{b} after it is propagated by \nf{R\sub{b}} but this would not be efficient because there might be $p$ subblocks in \nf{s}. However, once \nf{b} is installed, its superblock is going to be close to \nf{n.parent.head} at the time of installation. One idea is that if we know the aproximate position of the superblock of \nf{b} then we can search for the real superblock when we wished to know the superblock of \nf{b} i.e. \nf{b.super} does not have to be the exact location of  the superblock of \nf{b}, but we want it to be close to \nf{j}. We can set \nf{b.super} to \nf{n.parent.head} while creating \nf{b}, but the problem is that there might be many \nf{p.Refresh}es could happen that contain blocks from the right child of \nf{p} and \nf{j} could be arbitrarily (right word?) greater than \nf{b.super}. We set \nf{b.super} to \nf{p.head} after appending \nf{b} to \nf{n.blocks} (Line \ref{setSuper1}). Maybe \nf{R\sub{n}} goes to sleep at some time after installing \nf{b} and before setting \nf{b.super}. In this case the next \nf{Refresh}es on \nf{n} and \nf{n.parent} help fill in the value of \nf{b.super}.

Block \nf{b} is appended to \nf{n.blocks[h]} on Line \ref{cas}. After appending \nf{b}, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} by the same process or another process's \nf{n.Refresh()} or  maybe an \nf{n.parent.Refresh()}. We want to bound how far \nf{b.super} is from the index of \nf{b}'s superblock, which is created by a successful \nf{n.parent.Refresh()} that propagates \nf{b}.

\begin{observation} \label{setSupBeforeIncHead}
After \nf{n.blocks[i].CAS(null, b)} succeeds, \nf{n.head} cannot increase from \nf{i} to \nf{i+1} unless \nf{b.super} is set.
\end{observation}
\begin{proof}
From the Observation \ref{nonDecreasingHead} we know the only change to \nf{n.head} is on Line \ref{incrementHead} which is incrementing. Before an instance of \nf{Advance()} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
\end{proof}

\begin{corollary}
  If \nf{n.blocks[i].super} is \nf{null}, then $\nf{n.head}<\nf{i}$ and \nf{n.blocks[i+1]} is \nf{null}.
\end{corollary}
\begin{proof}If \nf{b.super} is \nf{null} then \nf{n.head} cannot advance so the next \nf{n.Refresh()} will fail. By the previous corollary, \nf{b.super} has to be set before the next successful Refresh() on n after \nf{R\sub{n}}.  
\end{proof}
Now let us talk about how the \nf{p.Refresh}es that took place after the putting \nf{b} into \nf{n}, will help to set \nf{b.super} and propagate \nf{b}.

\begin{lemma}
  If $\nf{b}\in\nf{n.parent.blocks[i]}$ then $\nf{b.super}\leq\nf{i}$.
\end{lemma}
\begin{proof}
For \nf{R\sub{p}} to contain block \nf{b}, it has to read \nf{n.head} greater than \nf{h} (see Line \ref{lastLine}). For \nf{n.head} to be greater than \nf{h} it means \nf{n.head} is incremented in Line \ref{incrementHead} which means \nf{b.super} was already set in Line \ref{setSuper1} (see Observation \ref{setSupBeforeIncHead}). So if \nf{R\sub{p}} propagates \nf{b} it means \nf{b.super} was already set. Let \nf{j} be the value written in \nf{b.super}. \nf{j} has been read in Line \ref{readHead} or Line \ref{readParentHead} which both are before calling \nf{Advance} that sets \nf{b.super}. From Observation \ref{nonDecreasingHead} we know \nf{p.Head} is non-decreasing so $\nf{j}\leq\nf{i}$. The reader may wonder when the case $\nf{j}=\nf{i}$ happens, it happens when \nf{p.blocks[j]=null} while \nf{j} is read and \nf{R\sub{p}} puts its created block into \nf{n.blocks[j]}.
\end{proof}

\begin{lemma}
If \nf{R\sub=n.Refresh()} puts \nf{b} in \nf{n.blocks[h]} at Line \ref{cas}, then the block created by one of the next two successful \nf{p.Refresh}es according to the Definition \ref{orderRefresh} contains \nf{b} and \nf{b.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{p.Refresh()}.
\end{lemma}

\begin{proof}
It is sufficient to prove one of the two successful \nf{p.Refresh()}es propagate \nf{b}. If the first successful \nf{p.Refresh()} propagated \nf{b} then the claim is true, so in the remaining part we assume the first \nf{p.Refresh()} did not propagate \nf{b} and prove the second \nf{p.Refresh()} propagates \nf{b}.

\nf{b.super} is set by some instance of \nf{Refresh()} on \nf{n} or \nf{p} showed by \nf{R'} and \nf{n.head} is incremented by some \nf{Refresh()} called \nf{R"}. We want to know how great $\nf{j}-\nf{b.super}$ can be. \nf{p.head} is \nf{hp} when \nf{R'} reads it. From Lemma \ref{lem::headProgress} \nf{p.head} could only increase from \nf{hp} to \nf{hp+1} if $\nf{p[hp]}\neq\nf{null}$. In other words there should be a successful \nf{p.Refresh()} for \nf{p.head} to increase. We claim there cannot be another successful \nf{p.Refresh()} after \nf{R'} reads \nf{p.head} and before \nf{R\sub{p}} performs Line \ref{lastLine}.

@@

Assume the first successful \nf{p.Refresh()} after $t_{\ref{cas}}^{Rn}$ did not set \nf{b.super}. It might happen maybe because the value read for $h_{left}$ in Line \ref{readChildHead} is less than \nf{i} or maybe $\nf{i}=\nf{h\sub{left}}$ and $\nf{left.blocks[h\sub{left}]}=\nf{null}$, which means \nf{n.head} is advanced but \nf{b} is still not installed in \nf{n.blocks[i]} which means \nf{R\sub{n}} has not reached to the Line \ref{cas}.

Let the first successful \nf{p.Refresh()} be $R_{p1}$ and the second next successful \nf{p.Refresh()} be $R_{p2}$. If $R_{p1}$ reads \nf{x} in Line \ref{readHead}, then $R_{p2}$ has to read \nf{x+1} in Line \ref{readHead} (induced from \ref{lem::headProgress}, \ref{nonDecreasingHead}). See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances $R_{p1}$,$R_{p2}$  on ${p}$.

\begin{figure}[hbt]  
  \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
  \caption{Rp2 reads p.head after $t_{321}^{Rp1}$, which is after $t_{321}^{Rn}$. Rp2 has to help increment n.head and set b.super.}
  \label{fig::timeLine2}
\end{figure}

$$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$$
$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}$ because $R_p$ is the first successful $p.Refresh()$ after $R_n$. $t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}$ because of the time windows between two consecutive successful \nf{Refresh()} instances on a node. $t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$ because it is sequential in the code. The second Refresh on \nf{p} contains \nf{b} and reads $\nf{n.head}>\nf{i}$.

So \nf{b.super} has set by some process before the second next successful \nf{p.Refresh()} on Line \ref{setSuper1}. Since \nf{i} is read in the Line \ref{readHead} then the \nf{CreateBlock()} in Line \ref{invokeCreateBlock} is going to read  some value fo \nf{left.head} greater than \nf{h} and propagates \nf{b} to \nf{p}. So if \nf{b} was not propagated already we are sure the second next successful \nf{p.Refresh()} propagates \nf{b}.
\end{proof}

\begin{corollary} \label{superRelation}
If \nf{$n$.blocks[$b$]} is propagated to $n.parent$, then \nf{$n$.blocks[$b$].super} is equal to or one less than the index of the superblock of \nf{$n$.blocks[$b$]}.
\end{corollary}

@@

So the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps. See Figure \ref{fig::doubling}. After Line \ref{dsearchEnd}, the binary search that finds \nf{b} also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$. Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}). So the claim is proved.
\end{proof}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/doubling.png}
  \caption{Distance relations between \nf{start}$,b,end$.}
  \label{fig::doubling}
\end{figure}

@@

>>>>

\pagebreak
\section{Our Queue}
Jayanti and Petrovic introduced a wait-free polylogarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We benefit from some ideas of their work to design a a polylogarithmic multi-enqueuer multi-dequeuer queue. Our algorithm despite them does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent operations. I our model there are $p$ procesess doing \nf{Enqueue(),Dequeue()} operations concurrently. We use a shared tree among the processes (see Figure~\ref{fig::blocktree}) to agree on one total ordering on the operations invoked by processes. Each process has a leaf which the order of operations invoked by the process is stored in it. When a process wishes to do an operation it appends the operation to its leaf and then tries to propagate its new operation up to the tree's root. In each node the ordering of operations propagated up to it is stored. All processes agree on the sequence stored in the root and it is defined to be the linearization ordering. 
\begin{figure}[hbt]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption{\label{fig::blocktree}Each process has a leaf and in each node there is an ordering of operations stored. Each node tries to propagate its operations up to the root, which stores the total ordering of all operations.}  
\end{figure}

@@

 Let $i$ be the value $R_n$, a successful instance of \nf{Refresh()} on node \nf{n} reads from \nf{$n$.head}. $R_n$ does a successful \nf{CAS(null, $b$)} into \nf{n.blocks[$i$]}. Let $p$ be $n$\nf{}.parent. Without loss of generality for the rest of this section assume ${n}$ is the left child of $p$. From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh()} propagating $b$. Let $R_p$ be the first successful \nf{p.Refresh()} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}. Let the index of the block $R_p$ put in $p$\nf{.blocks} be $j$.

Since the index of the superblock of ${b}$ is not known until $b$ is propagated, $R_n$ cannot set the \nf{super} field of ${b}$ while creating it. One approach is to set the \nf{super} field of ${b}$ by $R_n$ after propagating $b$ to $p$. This solution would not be efficient because there might be $p$ subblocks in the block $R_p$ propagated needing to update the \nf{super} field. However intuitively, once $b$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation. One idea is that if we know the approximate position of the superblock of $b$ then we can search for the real superblock when we wished to know the superblock of $b$ i.e. \nf{b.super} does not have to be the exact location of  the superblock of $b$, but we want it to be close to \nf{j}. We can set \nf{$b$.super} to \nf{$n$.parent.head} while creating $b$, but the problem is that there might be many \nf{$p$.Refresh}es that could happen after reading $p$\nf{.head} by $R_n$ and before propagating $b$ to $p$. If we set \nf{$b$.super} to \nf{$p$.head} after appending $b$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $b$ and before setting \nf{$b$.super}. In this case the next \nf{Refresh}es on $n$ and \nf{n.parent} help fill in the value of \nf{$b$.super}.

Block $b$ is appended to \nf{n.blocks[h]} on Line \ref{cas}. After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh()} by the same process or another process or  maybe an \nf{$n$.parent.Refresh()}. We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh()} that propagates $b$.

%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
\paragraph{Queue from tree}
Now, we describe how to use the tree to implement a queue. Consider the following execution of operations. \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. What should a \texttt{Dequeue()} return? To compute the response of a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.

@@

The next lemma is used to prove the subblocks of two blocks in a node are disjoint.
\begin{lemma} \label{lem::headProgress}
 If $n\nf{.blocks[}i\nf{]}\neq\nf{null}$ and $i>0$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
\end{lemma}
\begin{proof}
Consider the block $b$ written into \nf{n.blocks[$i$]} by \nf{CAS} at Line \ref{cas}. Block $b$ is created by the \nf{CreateBlock($i$)} called at Line \ref{invokeCreateBlock}. Prior to this call to \nf{CreateBlock($i$)}, $\nf{n.head}=i$ at Line \ref{readHead}, so \nf{n.blocks[$i-1$]} is already a non-null value $b$$^\prime$ by Invariant \ref{lem::headPosition}. Thus, the \nf{CreateBlock($i-1$)} that created  $b^\prime$ terminated before the \nf{CreateBlock($i$)} that creates $b$ is invoked. The value written into \nf{$b$.end\sub{left}} at Line \ref{endDefLine} of \nf{CreateBlock($i$)}  was one less than the value read at Line \ref{lastLine} of \nf{CreateBlock($i$)}. Similarly, the value in \nf{n.blocks[$i-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($i-1$)}. By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $b^\prime\nf{.end\sub{left}} \leq \nf{$b$.end\sub{left}}$. The proof for \nf{end\sub{right}} is similar.
\end{proof}

@@

Now we can define the operations of a block using the definition of subblocks.
\begin{definition}[Operations of a block]\label{def::ops}
A block $b$ in a leaf represents an \nf{Enqueue()} if $b\nf{.element}\neq\nf{null}$. Otherwise, if $b\nf{.element}=\nf{null}$, $b$ represents a \nf{Dequeue()}. The set of operations of block $b$ is the union of the operations in leaf subblocks of $b$. We denote the set of operations of block $b$ by ${ops(b)}$ and the union  of operations of a set of blocks $B$ by $ops(B)$.  We also say ${b}$ contains ${op}$ if $op\in{ops(b)}$.
\end{definition}

@@

\subsection{Correctness of GetEnqueue}

Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnqueue()}.

@@

For the induction step we prove if \nf{$n$.child.GetEnqueue($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$.  From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}). \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}. By Definitions \ref{def::ops}, \ref{def::subblock} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
\begin{align*}
&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \cup \\
&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
 \end{align*}
First we consider the case where the \nf{Enqueue} we are looking is in the left child. There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

@@

\clearpage

\subsection{Correctness of IndexDequeue}
\begin{definition}\label{orderRefresh} If \nf{Refresh} instance $R_1$ does its \nf{CAS} at Line \ref{cas} earlier than \nf{Refresh} instance $R_2$ we say $R_1$ has \it{happened before} $R_2$.
\end{definition}

\begin{observation} \label{setSupBeforeIncHead}
After \nf{$n$.blocks[$i$].CAS(null, $b$)} succeeds, \nf{$n$.head} cannot increase from ${i}$ to ${i+1}$ unless \nf{$b$.super} is set.
\end{observation}
\begin{proof}
From the Observation \ref{nonDecreasingHead} we know the only change to \nf{n.head} is on Line \ref{incrementHead} which is incrementing. Before an instance of \nf{Advance()} increments \nf{n.head} on Line \ref{incrementHead}, Line \ref{setSuper1} ensures that \nf{n.blocks[head].super} was set at Line \ref{setSuper1}.
\end{proof}

\begin{corollary}
  If \nf{n.blocks[i].super} is \nf{null}, then $\nf{n.head}<\nf{i}$ and \nf{n.blocks[i+1]} is \nf{null}.
\end{corollary}
\begin{proof}
By Lemma \ref{lem::headPosition} and Observation \ref{setSupBeforeIncHead}.
\end{proof}
Now let us talk about how the \nf{$p$.Refresh}es that took place after the putting $b$ into \nf{n}, will help to set \nf{b.super} and propagate $b$.

\begin{lemma}
  If $b$ is a direct subblock of \nf{$n$.parent.blocks[$i$]} then $\nf{$b$.super}\leq i$.
\end{lemma}
\begin{proof}
Instance $R_p$ of $n$\nf{.parent.Refresh()} contains block $b$, if it reads \nf{$n$.head} greater than ${h}$, the index of $b$ in $n$ (see Line \ref{lastLine}). For \nf{$n$.head} to be greater than ${h}$ it means \nf{$n$.head} is incremented in Line \ref{incrementHead} which means \nf{$b$.super} was already set in Line \ref{setSuper1} (see Observation \ref{setSupBeforeIncHead}). So if $R_p$ propagates $b$ it means \nf{$b$.super} was already set. The value written in \nf{b.super} has been read in Line \ref{readHead} or Line \ref{readParentHead} which both are before calling \nf{Advance} that sets \nf{$b$.super}. From Observation \ref{nonDecreasingHead} we know \nf{$n$.parent.head} is non-decreasing so $b\nf{.super}\leq\nf{i}$. The reader may wonder when the case $b\nf{.super}=\nf{i}$ happens, it happens when $\nf{$n$.parent.blocks[$b$.super]}=\nf{null}$ while $b$\nf{.super} is read and $R_p$ puts its created block into \nf{n.blocks[$b$\nf{.super}]}.
\end{proof}

\begin{lemma}
If $R_n=$\nf{Refresh()} puts $b$ in \nf{$n$.blocks[$h$]} at Line \ref{cas}, then the block created by one of the next two successful \nf{$n$.parent.Refresh}es according to the Definition \ref{orderRefresh} contains $b$ and \nf{$b$.super} is set before Line \ref{invokeCreateBlock} of the the second successful \nf{$n$.parent.Refresh()}.
\end{lemma}

\begin{proof}
It is sufficient to prove one of the two successful \nf{$n$.parent.Refresh()}es propagates $b$. If the first successful \nf{p.Refresh()} propagated $b$ then the claim is true. We prove if the first \nf{$n$.parent.Refresh()} did not propagate $b$, second \nf{$n$.parent.Refresh()} propagates $b$.

\nf{$b$.super} is set by some instance of \nf{Refresh()} on $n$ or \nf{$n$.parent} showed by $R^\prime$ and \nf{$n$.head} is incremented by some \nf{Refresh()} called $R^\prime\prime$. Let \nf{$n$.parent.head} value be $hp$ when $R^\prime$ reads it. From Lemma \ref{lem::headProgress} \nf{$n$.parent.head} could only increase from ${hp}$ to ${hp+1}$ if $\nf{$n$.parent.blocks[$hp$]}\neq\nf{null}$. In other words there should be a successful \nf{$n$.parent.Refresh()} for \nf{$n$.parent.head} to be increased. Let $R_p$ be \nf{$n$.parent.Refresh} that propagated $b$ to \nf{$n$.parent}. We claim there cannot be another successful \nf{$n$.parent.Refresh()} after $R^\prime$ reads \nf{$n$.parent.head} and before $R_p$ performs Line \ref{lastLine}.

@@

Without loss of generality  assume $n$ is the left child of \nf{$n$.parent}. As we mentioned the first successful \nf{$n$.parent.Refresh()} after $t_{\ref{cas}}^{Rn}$ did not set \nf{$b$.super}. It might happen maybe because the value $h_{left}$ read for $n$\nf{.head} in Line \ref{readChildHead} is less than or equal to $i$ and $\nf{left.blocks[$h_{left}$]}=\nf{null}$, which means \nf{$n$.head} is advanced but $b$ is still not installed in \nf{$n$.blocks[$i$]} which means $R_n$ has not reached  the Line \ref{cas}.

Let the first successful \nf{$n$.parent.Refresh()} be $R_{p1}$ and the second next successful \nf{$n$.parent.Refresh()} be $R_{p2}$. If $R_{p1}$ reads $x$ in Line \ref{readHead}, then $R_{p2}$ has to read ${x+1}$ in Line \ref{readHead} (induced from \ref{lem::headProgress}, \ref{nonDecreasingHead}). See the timeline in Figure \ref{fig::timeLine2} for two consecutive successful \nf{Refresh()} instances $R_{p1}$,$R_{p2}$ on \nf{$n$.parent}.

\begin{figure}[hbt]  
  \center\includegraphics[width=6in,height=1.25in]{pics/timeLineSuper2.png}
  \caption{$R_{p2}$ reads \nf{$n$.parent.head} after $t_{320}^{R_{p1}}$, which is after $t_{320}^{Rn}$. $R_{p2}$ has to help increment \nf{$n$.head} and set $b$\nf{.super}.}
  \label{fig::timeLine2}
\end{figure}

$$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$$
$t_{\ref{cas}}^{R_n}<t_{\ref{cas}}$ because $R_p$ is the first successful $p.Refresh()$ after $R_n$. $t_{\ref{cas}}^{R_p}<t_{\ref{cas}}^{R_p'}$ because of the time windows between two consecutive successful \nf{Refresh()} instances on a node. $t_{\ref{cas}}^{R_p'}<t_{314}^{R_p'}$ because it is sequential in the code. The second Refresh on \nf{p} contains $b$ and reads $\nf{n.head}>\nf{i}$.

So \nf{$b$.super} has set by some process before the second next successful \nf{$n$.parent.Refresh()} on Line \ref{setSuper1}. Since $i$ is read in the Line \ref{readHead} then the \nf{CreateBlock()} in Line \ref{invokeCreateBlock} is going to read  some value for \nf{left.head} greater than ${h}$ and propagates $b$ to \nf{$n$.parent}. So if $b$ was not propagated already we are sure the second next successful \nf{$n$.parent.Refresh()} propagates $b$.
\end{proof}

\begin{corollary} \label{superRelation}
If \nf{$n$.blocks[$b$]} is propagated to $n$\nf{.parent}, then \nf{$n$.blocks[$b$].super} is equal to or one less than the index of the superblock of \nf{$n$.blocks[$b$]}.
\end{corollary}

@@

So the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps. See Figure \ref{fig::doubling}. After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$. Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}). So the claim is proved.
\end{proof}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/doubling.png}
  \caption{Distance relations between \nf{start}$,b,end$.}
  \label{fig::doubling}
\end{figure}

@@

\section*{References}
\bibliography{main.bib} 
\bibliographystyle{plain}
